High geomagnetic field intensity recorded by anorthosite xenoliths requires a vigorous late Mesoproterozoic geodynamo

### 2022-01-25 PNAS_manuscript/PNAS_manuscript.tex

% Please add a significance statement to explain the relevance of your work
\significancestatement{Acquiring high-fidelity observational records of ancient magnetic field intensity from the remanent magnetization of rocks is crucial for constraining the evolution of Earth's core. However, robust estimates of ancient field strengths are often difficult to recover due to alteration or non-ideal rock magnetic behavior. In this study, we use plagioclase cumulates (anorthosite) that formed in the deep crust and were brought to the near surface by magma where they cooled and acquired thermal remanent magnetizations. These anorthosite xenoliths from the North American Midcontinent Rift have experienced minimal alteration since their emplacement and yield high quality paleointensity estimates. These data record a strong present-day-like virtual dipole moment 1.1 billion years ago. A strong field persisted over a 25 million year or greater interval indicating the existence of appreciable power sources for Earth's dynamo during the late Mesoproterozoic. }

@@

% At least three keywords are required at submission. Please provide three to five keywords, separated by the pipe symbol.
\keywords{absolute paleointensity $|$ Laurentia $|$ Midcontinent Rift $|$ anorthosite $|$ geodynamo $|$ Mesoproterozoic $|$ inner core}

\begin{abstract}
Obtaining estimates of Earth's magnetic field strength in deep time is complicated by non-ideal rock magnetic behavior in many igneous rocks. In this study, we target anorthosite xenoliths that cooled and acquired their magnetization within \textit{ca.} 1092 Ma shallowly emplaced diabase intrusions of the North American Midcontinent Rift. In contrast to the diabase which fails to provide reliable paleointensity estimates, the anorthosite xenoliths are unusually high-fidelity recorders yielding high-quality, single-slope paleointensity results that are consistent at specimen and site levels. An average value of $\sim$89 ZAm$^2$ for the virtual dipole moment from the anorthosites is higher than that of the dipole component of Earth's magnetic field today. Such a strong late Mesoproterozoic dipole recorded by the anorthosite xenoliths requires the existence of a vigorous geodynamo at the time. Together with previous paleointensity data from other Midcontinent Rift rocks, our new results indicate that a dynamo with strong power sources persisted for at least $\sim$25 million years \textit{ca.} 1.1 Ga. These data are inconsistent with there being a progressive monotonic decay of Earth's dynamo strength through the Proterozoic Eon.
\end{abstract}

@@

% If your first paragraph (i.e. with the \dropcap) contains a list environment (quote, quotation, theorem, definition, enumerate, itemize...), the line after the list may have some extra indentation. If this is the case, add \parshape=0 to the end of the list environment.
\dropcap{E}arth's magnetic field is the result of convective flow of liquid iron-alloy in Earth's outer core. At present day, the geodynamo is collectively powered by heat flow across the core-mantle boundary (CMB), and from the crystallization of the solid inner core from the liquid outer core which provides latent heat and compositional buoyancy. However, while paleomagnetic studies have found that a dynamo field has existed since at least 3.4 billion years ago \cite{Selkin2007a, Tarduno2014a, Brenner2020a}, Earth's inner core likely crystallized more recently. Estimates of the timing of the initial crystallization of Earth's inner core are interconnected with estimates for the core's thermal conductivity. Higher conductivity values imply faster cooling rates, which in turn imply that the threshold for the freezing of the inner core happened more recently \cite{Davies2015a}. While some estimates of these values are consistent with an inner core age $>$3 Ga \cite{Gubbins2004a, Konopkova2016a}, other estimates have implied higher thermal conductivity values and an age for the inner core that is less than 1.3 Ga \cite{Pozzo2012a, Koker2012a, Gomi2013a, Zhang2020b}, with some suggesting even younger ages ($<$ 700 Ma; ref. \citealp{Labrosse2015a, Ohta2016a}). Given that estimates for the core's thermal conductivity continue to be debated, it is crucial to use observational records as an independent constraint on the thermal evolution of Earth's core and mantle.

Paleomagnetic records from ancient rocks are one of the few types of observational data that have the potential to record the thermal evolution of Earth’s core. Evidence for a persistent magnetic field through the Proterozoic, for example, necessitates the existence of plate tectonics that sustained core-mantle boundary heat flow \cite{Swanson-Hysell2021c}. However, strikingly low estimates of geomagnetic field strengths have been obtained in the Ediacaran Period \cite{Bono2019a, Shcherbakova2019a,Thallner2021b} and the Devonian Period \cite{Shcherbakova2017a, Shcherbakova2021a}, potentially indicating unusual periods of core dynamo regime at those times. 
% These data have been interpreted to indicate that there was a progressively decaying field up to that time that was followed soon after by initial crystallization of the inner core \cite{Bono2019a}.

@@

Determinations of the absolute value of the the ancient geomagnetic field strength rely on igneous rocks that acquire thermal remanent magnetizations as they cool. These magnetizations need to be unmodified by subsequent heating or chemical alteration in order to maintain a record of the ancient geomagnetic field from the time of cooling. Intracontinental magmatic events are therefore an important target for determination of ancient paleointensity as they can be well-preserved within continental interiors. This interior position results in them typically being distant from tectonic events along continental margins that can drive alteration through heat and fluids. However, large igneous provinces that are often associated with intraplate magmatism are typically of geologically short duration with the bulk of magmatic products emplaced within 1 Myr or less \cite{Kasbohm2021a}. The Midcontinent Rift is an exception as it is a large igneous province where magmatism lasted $\sim$25 Myr from \texit{ca.} 1109 Ma to 1084 Ma \cite{Swanson-Hysell2019a}. Additionally, extension ceased in the Midcontinent Rift prior to lithospheric separation, preserving volcanic, intrusive, and sedimentary rocks of the rift within the continental interior. As a result, rocks of the rift have unusually simple paleomagnetic behavior for their greater than one billion-year-old age and form a central record of Mesoproterozoic paleogeography \cite{Swanson-Hysell2021c}. The duration of magmatic activity within the Midcontinent Rift is longer than the entire 20.4 Myr long Neogene Period such that it enables an extended well-preserved window into the intensity of Earth's magnetic field in the late Mesoproterozoic.

Despite the excellent preservation of the rocks, non-ideal paleointensity behaviors and overall low success rates have often led to high failure rates and challenged the interpretation of many previous paleointensity results from the Midcontinent Rift \cite{Pesonen1983a, Kulakov2013a}. The most trusted type of paleointensity estimate is that obtained through experiments in which the primary natural thermal remanent magnetization is progressively replaced by a laboratory magnetization that is imparted in a known field with internal consistency checks (such as in the IZZI-style Thellier experiments; ref. \citealp{Yu2004a}). In such Thellier paleointensity experiments, one typical departure from ideal behavior is sagging or double-slopes as visualized in plots that show thermal remanent magnetization (TRM) acquired versus natural remanent magnetization (NRM) lost. For such data, distinct paleointensity estimates may be calculated depending on the interpreter's choice of slope. Typically, such non-ideal behavior would result in a higher paleointensity estimate from the steeper-sloped low-temperature portion and a lower paleointensity estimate from the high-temperature portion. For example, in data from the Midcontinent Rift, ref. \citealp{Pesonen1983a} used the low-temperature slope as the best representation of the past magnetic field strength (likely overestimating the field strength) whereas ref. \citealp{Kulakov2013a} used the high-temperature slope (likely underestimating the field strength). Results with such non-ideal behavior were rejected by ref. \citealp{Sprain2018a} who applied stricter paleointensity selection criteria, but as a result had few accepted sites.

In this study, we target a unique rock type---high-purity anorthosite xenoliths. Anorthosites are cumulate rocks composed dominantly of the mineral plagioclase. They are attractive targets for paleomagnetic study as plagioclase crystals can enshroud minute magnetic inclusions from alteration with alteration of the plagioclase crystals not resulting in the formation of secondary iron oxides. The anorthosite xenoliths targeted in this study were brought to the near surface in magma that formed hypabyssal intrusions of the Beaver River diabase \cite{Zhang2021b}. They cooled and acquired their magnetizations in conjunction with the host diabase. Paleointensity experiments on the anorthosite xenoliths have a very high success rate, yielding consistent specimen- and site-level paleointensity results. Rock magnetic data reveal that the anorthosite xenoliths have low anisotropy of thermal remanent magnetization (TRM) and can acquire TRM linearly within relevant field strengths. Magnetic imaging shows the anorthosite specimens have dominant magnetic carriers interstitial to and within plagioclase crystals and they do not display strong preferred orientations. Our single-slope, high quality paleointensity data add to the reliable paleointensity records during $\sim$25 Myr of Midcontinent Rift magmatic activity and require that there was a strong late Mesoproterozoic geodynamo.

@@

%\begin{figure*}
\begin{SCfigure*}
\centering
\noindent\includegraphics[width=10.5 cm]{Petro_QDM.png}
\caption{\footnotesize{Thin section petrographic and magnetic images of anorthosites from the Beaver River anorthosite xenolith from which paleomagnetic site AX16 and geochronology sample MS99033 were collected (left column; ref. \citealp{Zhang2021b}) and distinct anorthosite from the Duluth Complex Anorthositic Series (right column). The Duluth Complex anorthosites were not targeted for paleointensity experiments in this study given the complexities associated with more pronounced fabrics. Cross-polarized petrographic images (A) Beaver River anorthosite have plagioclase with a granoblastic texture; (B) Plagioclase of Duluth Complex anorthosite have euhedral, interlocking crystals showing igneous foliation; (C) zoomed-in image of plagioclase from MS99033 showing the crystals to be largely free of large opaque inclusions; (D) Duluth Complex plagioclase can contain abundant Fe-Ti oxide needles that have preferred orientations that are often parallel with the [010] planes of the plagioclase crystals. Magnetic field maps acquired by the quantum diamond microscope (QDM) show that the relatively weak magnetic sources within and interstitial to plagioclase crystals exist in the Beaver River anorthosite in (E) while large oxide needles within Duluth Complex plagioclase are strongly magnetic as evidenced through the overlapping magnetic fields above the oxides in (F). Magnetic field maps of (E) and (F) share the same spatial and magnetic scale. The unit of the axes are in $\mu$m.}}
\label{fig:Petro_QDM}
%\end{figure*}
\end{SCfigure*}

\section*{Results and Discussion}

@@

Petrographic imaging shows the dominantly monomineralic anorthosite xenoliths within the Beaver River diabase often have granoblastic texture characterized by equigranular crystals with weakly developed petrofabrics (Fig. \ref{fig:Petro_QDM}A,C). In addition, the plagioclase crystals of the anorthosite xenoliths are typically free of large oxide inclusions. Magnetic imaging using the quantum diamond microscope (QDM) shows that the dominant magnetic remanence carriers often exist within, and sometimes interstitial to, plagioclase crystals (Fig. \ref{fig:Petro_QDM}E).

These features of the Beaver River anorthosite xenoliths are distinct from other anorthosites such as the anorthosite xenoliths within the Duluth Complex Anorthositic Series rocks---older intrusions within the Midcontinent Rift. To illustrate these difference, we present petrographic and magnetic imaging data from a sample of Duluth Complex anorthosite that was not targeted for paleointensity experiments. The plagioclase of the Duluth Complex anorthosites typically develop interlocking textures that display strong igneous foliation (Fig. \ref{fig:Petro_QDM}B). In addition, there are abundant Fe-Ti oxide needles that are typically aligned with the [010] planes of plagioclase crystals (Fig. \ref{fig:Petro_QDM}D). Magnetic imaging confirms that those strongly magnetic needles within the plagioclase crystals often have magnetic moments oriented along their long axes (Fig. \ref{fig:Petro_QDM}D,F). Titanomagnetite and ilmenite symplectic intergrowths also coexist with pyroxene and relict olivine as a product of olivine oxidation in the Duluth Complex plagioclase \cite{Miller1990a}.

@@

Following IZZI-modified paleointensity experiments \cite{Yu2004a}, 40 from a total of 86 anorthosite specimens and 0 out of a total of 69 diabase specimens passed our paleointensity result selection criteria (see Materials and Methods section). 7 anorthosite sites and no diabase site-level results pass these selection criteria. Example NRM/TRM (Arai) plots are shown in Fig. \ref{fig:IZZI_examples}. Summary specimen absolute paleointensity estimates and site-level weighted mean paleointensity values are plotted in Fig. \ref{fig:PINT_cooling_corrected} (and provided in Table S1). The paleointensity quality index (Q$_{PI}$; ref. \citealp{Biggin2014a}) for the anorthosite xenoliths are all above 4 (Table S2). The cooling rate-corrected absolute paleointensity estimates from all anorthosite sites have a weighted mean of 52.72 $\pm $5.72 $\mu$T. This corresponds to a calculated virtual dipole moment of $\sim$89 ZAm$^2$ ($10^{21} Am^2$) \textit{ca.} 1092 Ma. All measurement-level paleointensity experiment data is available within the MagIC database (\url{}; ATTACHED IS A LINK FOR REVIEWING PURPOSESS, THE URL WILL BE UPDATED WHEN DOI IS GENERATED).

@@

% Paleointensity experiment results from the anorthosite xenoliths and their host diabase show three general types of dominant specimen behaviors on NRM/TRM plots (Fig. \ref{fig:IZZI_examples}): dominantly single-slope, double-slope or sagging, and zigzagging. 
Typical paleointensity experimental data of the anorthosite specimens have straight, single-slope NRM/TRM plots and the accepted fractions of temperature steps span over the origin-trending, primary remanence components (Fig. \ref{fig:IZZI_examples}). We accept specimen- and site-level absolute paleointensity results from those anorthosite xenoliths that pass the selection criteria. Other anorthosite xenoliths and diabase specimens failed the selection criteria largely because of double-slope or sagging behavior (fail FRAC selection; see methods), poor pTRM checks, and sometimes zigzagging behaviors superimposed on top (fail SCAT, DANG selection; see methods; Fig. \ref{fig:IZZI_examples}). A 20 mT alternating field treatment after in-field heating steps was applied to some specimens, but it did not result in significant changes in the experimental results for the anorthosite xenolith or diabase specimens (Fig. \ref{fig:PINT_cooling_corrected}; Supporting Information).

@@

Rock magnetic data support that the anorthosite specimens that pass the paleointensity selection criteria have dominant magnetic remanence carriers with magnetic properties similar to stochiometric, non-interacting, single domain magnetite, whereas anorthosite samples that failed the paleointensity result selection and all diabase samples have more pronounced populations of non-ideal carriers. Magnetic property measurement system (MPMS) data show both diabase and anorthosite contain (titano)magnetite as evidenced through the presence of the Verwey transition (Fig. S2; ref. \citealp{Verwey1939a, Feinberg2015a}). Anorthosite chips from sites that yield successful paleointensity results have Verwey transition temperatures near 120 Kelvin (K) as expected for stochiometric magnetite with minimal Ti \cite{Ozdemir1993a}. However, diabase and anorthosite specimens that did not pass our paleointensity selection typically have Verwey transitions that are suppressed toward lower temperatures (Fig. S2), indicating that the specimens either has a relatively higher Ti content or has been partially oxidized \cite{Ozdemir1993a}.

Coercivity spectra from backfield experiments further support the interpretation that rocks that pass or fail the selection criteria have distinct magnetic mineralogy. A compilation of median destructive field (MDF) values in Fig. \ref{fig:coercivity} shows that specimens from anorthosites that pass paleointensity selections have a distinctly higher average MDF values ($\sim$60 mT) than other anorthosites and diabase ($\sim$30 mT). Single-component fits for coercivity spectra \cite{Maxbauer2016a} show that anorthosites that yielded successful paleointensity results can have magnetic grain populations with peak coercivity around 80 mT (Fig. \ref{fig:coercivity}). In contrast, other anorthosite and diabase tend to have lower coercivities ($\sim$30 mT). This result is consistent with the interpretation that they lack a dominance of single domain magnetite grains in their magnetic mineral assemblages and thus have non-ideal paleointensity behaviors.

@@

\subsection*{TRM anisotropy and nonlinear acquisition check}
Significant remanence anisotropy has been documented to exist within certain anorthositic rocks that form in layered intrusive complexes \cite{Selkin2000a, Feinberg2006a}. Strong remanence anisotropy associated with the igneous foliation developed within anorthosite from the Stillwater Complex has been shown to lead to significant overestimate or underestimate of paleointensity values depending on the relative orientations between the fabrics and an applied magnetic field \cite{Selkin2000a}. To assess whether our paleointensity estimates are biased by remanence anisotropy, we calculated the gamma statistic, which is the angular difference between the last pTRM step of paleointensity experiment and the applied field direction. The results show that the anorthosite specimens used for paleointensity experiment have low gamma values ranging from $\sim$0.9\textdegree to $\sim$11.9\textdegree, with a median value of 4.2\textdegree (Table S1). These gamma values are similar to those of Midcontinent Rift volcanics \cite{Sprain2018a}. Therefore, the Beaver River anorthosite xenolith bulk samples do not have a significant remanence anisotropy. Paleodirectional data from our anorthosite xenoliths further support that they have minimal remanence anisotropy as their site mean directions closely match those of the Beaver River diabase hosts without deviating due to a fabric \cite{Zhang2021b}. In addition, no resolvable magnetic fabric was observed in the anorthosite xenoliths through petrographic and magnetic imaging (Fig. \ref{fig:Petro_QDM}).

We applied full TRMs to a set of our anorthosite specimens that pass the paleointensity selection criteria with lab fields of 30, 50, 70, and 90 $\mu$T along specimen vertical axes. The results show that our anorthosite specimens did not saturate upon the applied fields and there is thus no need for correcting nonlinear thermal remanence acquisitions (ref. \citealp{Selkin2007a}; Fig. S4).

>>>>

% Please add a significance statement to explain the relevance of your work
\significancestatement{Acquiring high-fidelity observational records of ancient magnetic field intensity from the remanent magnetization of rocks is crucial for constraining the evolution of Earth's core. However, robust estimates of ancient field strengths are often difficult to recover due to alteration or non-ideal rock magnetic behavior. In this study, we use plagioclase cumulates (anorthosite) that formed in the deep crust and were brought to the near surface by magma where they cooled and acquired thermal remanent magnetizations. These anorthosite xenoliths from the North American Midcontinent Rift have experienced minimal alteration since their emplacement and yield high quality paleointensity estimates. These data record a strong present-day-like virtual dipole moment 1.1 billion years ago. A strong field persisted over a 25 million year or greater interval indicating the existence of appreciable power sources for Earth's dynamo during the late Mesoproterozoic.}

@@

% At least three keywords are required at submission. Please provide three to five keywords, separated by the pipe symbol.
\keywords{absolute paleointensity $|$ Midcontinent Rift $|$ anorthosite $|$ geodynamo $|$ Mesoproterozoic $|$ inner core}

\begin{abstract}
Obtaining estimates of Earth's magnetic field strength in deep time is complicated by non-ideal rock magnetic behavior in many igneous rocks. In this study, we target anorthosite xenoliths that cooled and acquired their magnetization within \textit{ca.} 1092 Ma shallowly emplaced diabase intrusions of the North American Midcontinent Rift. In contrast to the diabase which fails to provide reliable paleointensity estimates, the anorthosite xenoliths are unusually high-fidelity recorders yielding high-quality, single-slope paleointensity results that are consistent at specimen and site levels. An average value of $\sim$89 ZAm$^2$ for the virtual dipole moment from the anorthosites is higher than that of the dipole component of Earth's magnetic field today. Such a strong late Mesoproterozoic dipole recorded by the anorthosite xenoliths requires the existence of a vigorous geodynamo at the time. Together with previous paleointensity data from other Midcontinent Rift rocks, our new results indicate that a dynamo with strong power sources persisted for more than 20 million years \textit{ca.} 1.1 Ga. These data are inconsistent with there being a progressive monotonic decay of Earth's dynamo strength through the Proterozoic Eon.
\end{abstract}

@@

% If your first paragraph (i.e. with the \dropcap) contains a list environment (quote, quotation, theorem, definition, enumerate, itemize...), the line after the list may have some extra indentation. If this is the case, add \parshape=0 to the end of the list environment.
\dropcap{E}arth's magnetic field is the result of convective flow of liquid iron-alloy in Earth's outer core. At present day, the geodynamo is collectively powered by heat flow across the core-mantle boundary (CMB) and from the crystallization of the solid inner core from the liquid outer core which provides latent heat and compositional buoyancy \cite{Buffett2000a}. However, while paleomagnetic studies have found that a dynamo field has existed since at least 3.4 billion years ago \cite{Selkin2007a, Tarduno2014a, Brenner2020a}, Earth's inner core likely crystallized more recently. Estimates of the timing of the initial crystallization of Earth's inner core are interconnected with estimates for the core's thermal conductivity. Higher conductivity values imply faster cooling rates, which in turn imply that the threshold for the freezing of the inner core happened more recently \cite{Davies2015a}. While some estimates of these values are consistent with an inner core age $>$3 Ga \cite{Gubbins2004a, Konopkova2016a}, other estimates have implied higher thermal conductivity values and an age for the inner core that is less than 1.3 Ga \cite{Pozzo2012a, Koker2012a, Gomi2013a, Zhang2020b}, with some suggesting even younger ages ($<$ 700 Ma; ref. \citealp{Labrosse2015a, Ohta2016a}). Given that estimates for the core's thermal conductivity continue to be debated, it is crucial to use observational records as an independent constraint on the thermal evolution of Earth's core and mantle.

Paleomagnetic records from ancient rocks are one of the few types of observational data that have the potential to provide constraints on the thermal evolution of Earth’s core. Evidence for a persistent magnetic field through the Proterozoic, for example, necessitates the existence of plate tectonics that sustained core-mantle boundary heat flow \cite{Swanson-Hysell2021c}. However, strikingly low estimates of geomagnetic field strengths have been obtained during the 635 to 539 Ma Ediacaran Period \cite{Bono2019a, Shcherbakova2019a,Thallner2021b} and the 419 to 359 Ma Devonian Period \cite{Shcherbakova2017a, Shcherbakova2021a}, potentially indicating unusual periods of core dynamo regime at those times. The Ediacaran data have been interpreted to indicate that there was a progressively decaying field up to that time that was followed soon after by initial crystallization of the inner core \cite{Bono2019a}.

@@

Determinations of the absolute value of the ancient geomagnetic field strength rely on igneous rocks that acquire thermal remanent magnetizations as they cool. These magnetizations need to be unmodified by subsequent heating or chemical alteration in order to maintain a record of the ancient geomagnetic field from the time of cooling. Intracontinental magmatic events are therefore an important target for determination of ancient paleointensity as they can be well-preserved within continental interiors. This interior position results in them typically being distant from tectonic events along continental margins that can drive alteration through heat and fluids. However, intraplate magmatism associated with large igneous provinces is typically of geologically short duration with the bulk of magmatic products emplaced within 1 Myr or less \cite{Kasbohm2021a}. The Midcontinent Rift is an exception as it is a large igneous province where magmatism lasted $\sim$25 Myr from \texit{ca.} 1109 Ma to 1084 Ma \cite{Swanson-Hysell2019a}. Additionally, extension ceased in the Midcontinent Rift prior to lithospheric separation, preserving volcanic, intrusive, and sedimentary rocks of the rift within the continental interior. As a result, rocks of the rift have unusually simple paleomagnetic behavior for their greater than one billion-year-old age and paleomagnetic data from rift rocks forms a central record of Mesoproterozoic paleogeography \cite{Swanson-Hysell2021c}. The duration of magmatic activity within the Midcontinent Rift is longer than the entire 20.4 Myr long Neogene Period such that it enables an extended well-preserved window into the intensity of Earth's magnetic field in the late Mesoproterozoic.

Despite the excellent preservation of the rocks, non-ideal paleointensity behaviors have challenged the interpretation of many previous paleointensity results from the Midcontinent Rift \cite{Pesonen1983a, Kulakov2013a}. The most trusted type of paleointensity estimate is that obtained through experiments in which the primary natural thermal remanent magnetization is progressively replaced by a laboratory magnetization that is imparted in a known field with internal consistency checks (such as in IZZI-style Thellier experiments; ref. \citealp{Yu2004a}). In such Thellier paleointensity experiments, one typical departure from ideal behavior is sagging or double-slopes as visualized in plots that show thermal remanent magnetization (TRM) acquired versus natural remanent magnetization (NRM) lost. For such data, distinct paleointensity estimates may be calculated depending on the interpreter's choice of slope. Typically, such non-ideal behavior would result in a higher paleointensity estimate from the steeper-sloped low-temperature portion and a lower paleointensity estimate from the high-temperature portion. For example, in data from the Midcontinent Rift, ref. \citealp{Pesonen1983a} used the low-temperature slope as the best representation of the past magnetic field strength (likely overestimating the field strength) whereas ref. \citealp{Kulakov2013a} used the high-temperature slope (likely underestimating the field strength). Results with such non-ideal behavior were rejected by ref. \citealp{Sprain2018a} who applied stricter paleointensity selection criteria, but as a result had few accepted sites.

In this study, we target a unique rock type---high-purity anorthosite xenoliths. Anorthosites are cumulate rocks composed dominantly of the mineral plagioclase. They are attractive targets for paleomagnetic study as plagioclase crystals can enshroud minute magnetic inclusions from alteration with minor alteration of the plagioclase crystals not resulting in the formation of secondary iron oxides in contrast with mafic minerals such as olivine. The anorthosite xenoliths targeted in this study were brought to the near surface in magma that formed hypabyssal intrusions of the Beaver River diabase \cite{Zhang2021b}. They cooled and acquired their magnetizations in conjunction with the host diabase. Paleointensity experiments on the anorthosite xenoliths have a very high success rate, yielding consistent specimen- and site-level paleointensity results. Rock magnetic data reveal that the anorthosite xenoliths have low anisotropy of thermal remanent magnetization (TRM) and can acquire TRM linearly within relevant field strengths. Magnetic imaging shows the anorthosite specimens have dominant magnetic carriers interstitial to and within plagioclase crystals and they do not display strong preferred orientations. These single-slope, high quality paleointensity data, together with other high-quality paleointensity records during $\sim$25 Myr of Midcontinent Rift magmatic activity, require that there was a strong late Mesoproterozoic geodynamo.

@@

%\begin{figure*}
\begin{SCfigure*}
\centering
\noindent\includegraphics[width=13 cm]{Petro_QDM.png}
\caption{\footnotesize{Thin section petrographic images (A,B,C,D) and magnetic maps (E,F) of anorthosite from the Beaver River anorthosite xenolith from which paleomagnetic site AX16 and geochronology sample MS99033 were collected (left column; ref. \citealp{Zhang2021b}) and a distinct anorthosite from the Duluth Complex Anorthositic Series (right column). The Duluth Complex anorthosites were not targeted for paleointensity experiments in this study given the complexities associated with more pronounced fabrics. Cross-polarized petrographic images of the Beaver River anorthosite (A,C) reveal plagioclase with a granoblastic texture with platioclase crystals to are largely free of large opaque inclusions.  In contrast, plagioclase crystals in the Duluth Complex anorthosite have euhedral, interlocking crystals with an igneous foliation (B) and the plagioclase crystals contain abundant Fe-Ti oxide needles that have preferred orientations that are often parallel with the [010] planes of the crystals. Magnetic field maps acquired by the quantum diamond microscope (QDM) show that the relatively weak magnetic sources within and interstitial to plagioclase crystals exist in the Beaver River anorthosite in (E) while large oxide needles within Duluth Complex plagioclase are strongly magnetic as evidenced through the overlapping magnetic fields above the oxides in (F). Magnetic field maps of (E) and (F) share the same spatial and magnetic scale. The unit of the axes are in $\mu$m.}}
\label{fig:Petro_QDM}
%\end{figure*}
\end{SCfigure*}

\section*{Results and Interpretations}

@@

The dominantly monomineralic anorthosite xenoliths within the Beaver River diabase often have granoblastic texture characterized by equigranular crystals with weakly developed petrofabrics (Fig. \ref{fig:Petro_QDM}A,C). In addition, the plagioclase crystals of the anorthosite xenoliths are typically free of large oxide inclusions. Magnetic imaging using the quantum diamond microscope (QDM) shows that the dominant magnetic remanence carriers often exist within, and sometimes interstitial to, plagioclase crystals (Fig. \ref{fig:Petro_QDM}E).

These features of the Beaver River anorthosite xenoliths are distinct from other anorthosites such as the anorthosite xenoliths within the Duluth Complex Anorthositic Series rocks---older intrusions within the Midcontinent Rift. To illustrate these difference, we present petrographic and magnetic imaging data from a sample of Duluth Complex anorthosite that was not targeted for paleointensity experiments. The plagioclase of the Duluth Complex anorthosites typically develop interlocking textures that display strong igneous foliation (Fig. \ref{fig:Petro_QDM}B). In addition, there are abundant Fe-Ti oxide needles that are typically aligned with the [010] planes of the plagioclase crystals (Fig. \ref{fig:Petro_QDM}D). Magnetic imaging confirms that those strongly magnetic needles within the plagioclase crystals often have magnetic moments oriented along their long axes (Fig. \ref{fig:Petro_QDM}D,F). Titanomagnetite and ilmenite symplectic intergrowths also coexist with pyroxene and relict olivine as a product of olivine oxidation in the Duluth Complex plagioclase \cite{Miller1990a}.

@@

Following IZZI-modified paleointensity experiments \cite{Yu2004a}, 40 from a total of 86 anorthosite specimens and 0 out of a total of 69 diabase specimens passed our paleointensity result selection criteria (see Materials and Methods section). 7 anorthosite sites and no diabase site-level results pass these selection criteria. Example NRM/TRM (Arai) plots are shown in Fig. \ref{fig:IZZI_examples}. Summary specimen absolute paleointensity estimates and site-level weighted mean paleointensity values are plotted in Fig. \ref{fig:PINT_cooling_corrected} (and provided in Table S1). The paleointensity quality index (Q$_{PI}$; ref. \citealp{Biggin2014a}) for the anorthosite xenoliths are all above 4 (Table S2). The cooling rate-corrected absolute paleointensity estimates from all anorthosite sites have a weighted mean of 52.72 $\pm$ 5.72 $\mu$T. This corresponds to a calculated virtual dipole moment of $\sim$89 ZAm$^2$ ($10^{21} Am^2$) \textit{ca.} 1092 Ma. All measurement-level paleointensity experiment data is available within the MagIC database (\url{https://earthref.org/MagIC/19462/8d3c2258-11ae-4830-b99f-3f6b02eceb7e}; \textit{this private link is provided for the purpose of review; the url will be updated when a doi is generated for this manuscript}).

@@

Typical paleointensity experimental data of the anorthosite specimens have straight, single-slope NRM/TRM plots and the accepted fractions of temperature steps span over the origin-trending, primary remanence components (Fig. \ref{fig:IZZI_examples}). We accept specimen- and site-level absolute paleointensity results from those anorthosite xenoliths that pass the selection criteria. Other anorthosite xenoliths and diabase specimens failed the selection criteria largely because of double-slope or sagging behavior (fail FRAC selection; see methods), poor pTRM checks, and sometimes zigzagging behaviors superimposed on top of sagging behavior (fail SCAT, DANG selection; see methods; Fig. \ref{fig:IZZI_examples}). A 20 mT alternating field treatment after in-field heating steps was applied to some specimens, but it did not result in significant changes in the experimental results for the anorthosite xenolith or diabase specimens (Fig. \ref{fig:PINT_cooling_corrected}; Supporting Information).

@@

Rock magnetic data support that the anorthosite specimens that pass the paleointensity selection criteria have dominant magnetic remanence carriers with magnetic properties similar to stochiometric, non-interacting, single domain magnetite, whereas anorthosite samples that failed the paleointensity result selection and all diabase samples have more pronounced populations of non-ideal carriers. Magnetic property measurement system (MPMS) data show both diabase and anorthosite contain (titano)magnetite as evidenced through the presence of the Verwey transition (Fig. S2; ref. \citealp{Verwey1939a, Feinberg2015a}). Anorthosite chips from sites that yield successful paleointensity results have Verwey transition temperatures near 120 K as expected for stochiometric magnetite with minimal Ti \cite{Ozdemir1993a}. However, diabase and anorthosite specimens that did not pass our paleointensity selection typically have Verwey transitions that are suppressed toward lower temperatures (Fig. S2), indicating that the magnetite in the specimens either has relatively higher Ti content or has been partially oxidized \cite{Ozdemir1993a}.

Coercivity spectra from backfield experiments further support the interpretation that rocks that pass or fail the selection criteria have distinct magnetic mineralogy (Fig. \ref{fig:coercivity}). A compilation of median destructive field (MDF) values in Fig. \ref{fig:coercivity} shows that specimens from anorthosites that pass paleointensity selections have a distinctly higher average MDF values than other anorthosites and diabase. Single-component fits for coercivity spectra \cite{Maxbauer2016a} show that anorthosites that yielded successful paleointensity results can have magnetic grain populations with peak coercivity around 80 mT (Fig. \ref{fig:coercivity}). In contrast, other anorthosite and diabase tend to have lower peak coercivities ($\sim$30 mT). This result is consistent with an interpretation that a population of magnetic grains with more multidomain-like behavior is responsible for the non-ideal paleointensity behaviors during experiments on such specimens.

@@

\subsection*{TRM anisotropy and nonlinear acquisition check}
Significant remanence anisotropy has been documented to exist within certain anorthositic rocks that form in layered intrusive complexes \cite{Selkin2000a, Feinberg2006a}. Strong remanence anisotropy associated with the igneous foliation developed within anorthosite from the Stillwater Complex has been shown to lead to significant overestimate or underestimate of paleointensity values depending on the relative orientations between the fabrics and an applied magnetic field \cite{Selkin2000a}. To assess whether our paleointensity estimates are biased by remanence anisotropy, we calculated the gamma statistic, which is the angular difference between the last pTRM step of paleointensity experiment and the applied field direction. The results show that the anorthosite specimens used for paleointensity experiment have low gamma values ranging from $\sim$0.9\textdegree to $\sim$11.9\textdegree, with a median value of 4.2\textdegree$\;$(Table S1). These gamma values are similar to those of Midcontinent Rift volcanics \cite{Sprain2018a}. Therefore, the Beaver River anorthosite xenolith bulk samples do not have a significant remanence anisotropy. Paleodirectional data from our anorthosite xenoliths further support that they have minimal remanence anisotropy as their site mean directions closely match those of the Beaver River diabase hosts without deviating due to a fabric \cite{Zhang2021b}. In addition, no resolvable magnetic fabric was observed in the anorthosite xenoliths through petrographic and magnetic imaging (Fig. \ref{fig:Petro_QDM}).

We applied full TRMs to a set of anorthosite specimens with lab fields of 30, 50, 70, and 90 $\mu$T along specimen vertical axes. The results show that our anorthosite specimens did not saturate upon the applied fields and there is thus no need for correcting nonlinear thermal remanence acquisitions (ref. \citealp{Selkin2007a}; Fig. S4).

\subsection*{Averaging secular variations and applying cooling rate correction}
To best characterize the geomagnetic axial dipole field intensity during a certain time period a paleointensity dataset should cover a sufficient amount of time such that paleosecular variations of the geomagnetic field are averaged. Thermal modeling results from ref. \citealp{Zhang2021b} suggest that the Beaver River anorthosite xenoliths were likely heated to tholeiitic magma temperature ($\sim$1100\textdegree C) and acquired thermal remanent magnetization during cooling with their diabase host on a time scale of a thousand years, partially averaging secular variation within single samples. Anorthosite xenoliths have paleointensity results that are consistent within small regions, but vary between localities. Anorthosite AX12 and AX16 were emplaced $\sim$450 meters apart and have very similar paleointensity estimates. $\sim$10 km to the north, anorthosite xenoliths AX11 and AX13 were also emplaced closely at a distance of $\sim$125 meters and yield consistent and distinct paleointensity estimates from those of AX12 and AX16 (Fig. \ref{fig:Geologic_map}). The data therefore capture different intervals of time during the emplacement and cooling of the Beaver River diabase sills \textit{ca.} 1092 Ma.
